---
title: "Income inequality and housing affordabilities in The Netherlands"
author: "Toms Kelderis (2898504)"
date: "`r Sys.Date()`"
output: pdf_document
---
#Lester M.
# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")

```

```{r packages}
require(tidyverse)li
require(rmarkdown)
require(yaml)
library(dplyr)
```

#loading in data, not necessary to run it again
```{r}
quintile_income <- read.csv("data/83765NED_UntypedDataSet_04062026_151743.csv", sep = ";")

median_income <- read.csv("data/Inkomen_huishoudens__kenmerken__regio_03062026_161425.csv", sep = ";")

renter_households <- read.csv("data/Woningen__m___type__bouwjaar__regio_03062026_161143.csv", sep = ";")

#renaming quintile_income columns
quintile_income <- quintile_income %>%
  select(
    region_name        = WijkenEnBuurten,
    municipality       = Gemeentenaam_1,
    region_type        = SoortRegio_2,
    region_code        = Codering_3,
    avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
    pct_owner          = Koopwoningen_40,
    pct_renter         = HuurwoningenTotaal_41,
    pct_social_housing = InBezitWoningcorporatie_42,
    avg_income         = GemiddeldInkomenPerInwoner_66,
    pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
    pct_top20          = k_20PersonenMetHoogsteInkomen_68,
    pct_low_income     = HuishoudensMetEenLaagInkomen_72,
    pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73)

#filtering to municipalities only
quintile_income <- quintile_income %>%
  mutate(region_type = str_trim(region_type)) %>%
  filter(region_type == "Gemeente")

#fixing decimal commas
quintile_income <- quintile_income %>%
  mutate(across(c(avg_home_value, pct_owner, pct_renter, pct_social_housing,
                  avg_income, pct_bottom40, pct_top20, pct_low_income,
                  pct_poverty_line),
                ~ as.numeric(str_replace(., ",", "."))))

#renaming median_income columns
median_income <- median_income %>%
  select(
    population         = Populatie,
    household_type    = Kenmerken.van.huishoudens,
    year              = Perioden,
    region            = Regio.s,
    median_std_income = Inkomen.Mediaan.gestandaardiseerd.inkomen..1.000.euro.,
    median_disp_income = Inkomen.Mediaan.besteedbaar.inkomen..1.000.euro.)

#renaming renter_households columns
renter_households <- renter_households %>%
  select(
    housing_type  = Woningtype,
    year          = Perioden,
    region        = Regio.s,
    housing_stock = Beginstand.woningvoorraad..aantal.)

saveRDS(quintile_income, "data/quintile_income.rds")
saveRDS(median_income, "data/median_income.rds")
saveRDS(renter_households, "data/renter_households.rds")

#creating new variable
quintile_income <- quintile_income %>%
  mutate(
    affordability_ratio = avg_home_value / avg_income,
    inequality_gap      = pct_top20 - pct_bottom40)

new_variables <- quintile_income %>%
  select(municipality, avg_home_value, avg_income, affordability_ratio, inequality_gap)
```

```{r}
quantile_income <- readRDS("data/quintile_income.rds")
```

# Title Page

Toms Ķelderis (2898504), Nathan van Doorn (2801374), Frederick Nuboer (2809143), Khpelwaak Mamound (2786698), Caner Tepe (2785569), Mijke Gerritsen (2824547)

Tutorial 4, Group 1

Lecturer - Drs. Fatima Amankour

# Part 1 - Identify a Social Problem

Housing unaffordability and income inequality in the Netherlands

## 1.1 Describe the Social Problem

Text HERE

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
dataset <- midwest

quintile_income <- read.csv("data/83765NED_UntypedDataSet_04062026_151743.csv", sep = ";")

median_income <- read.csv("data/Inkomen_huishoudens__kenmerken__regio_03062026_161425.csv", sep = ";")

renter_households <- read.csv("data/Woningen__m___type__bouwjaar__regio_03062026_161143.csv", sep = ";")

#renaming quintile_income columns
quintile_income <- quintile_income %>%
  select(
    region_name        = WijkenEnBuurten,
    municipality       = Gemeentenaam_1,
    region_type        = SoortRegio_2,
    region_code        = Codering_3,
    avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
    pct_owner          = Koopwoningen_40,
    pct_renter         = HuurwoningenTotaal_41,
    pct_social_housing = InBezitWoningcorporatie_42,
    avg_income         = GemiddeldInkomenPerInwoner_66,
    pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
    pct_top20          = k_20PersonenMetHoogsteInkomen_68,
    pct_low_income     = HuishoudensMetEenLaagInkomen_72,
    pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73)

#filtering to municipalities only
quintile_income <- quintile_income %>%
  mutate(region_type = str_trim(region_type)) %>%
  filter(region_type == "Gemeente")

#fixing decimal commas
quintile_income <- quintile_income %>%
  mutate(across(c(avg_home_value, pct_owner, pct_renter, pct_social_housing,
                  avg_income, pct_bottom40, pct_top20, pct_low_income,
                  pct_poverty_line),
                ~ as.numeric(str_replace(., ",", "."))))

#renaming median_income columns
median_income <- median_income %>%
  select(
    population         = Populatie,
    household_type    = Kenmerken.van.huishoudens,
    year              = Perioden,
    region            = Regio.s,
    median_std_income = Inkomen.Mediaan.gestandaardiseerd.inkomen..1.000.euro.,
    median_disp_income = Inkomen.Mediaan.besteedbaar.inkomen..1.000.euro.)

#renaming renter_households columns
renter_households <- renter_households %>%
  select(
    housing_type  = Woningtype,
    year          = Perioden,
    region        = Regio.s,
    housing_stock = Beginstand.woningvoorraad..aantal.)

saveRDS(quintile_income, "data/quintile_income.rds")
saveRDS(median_income, "data/median_income.rds")
saveRDS(renter_households, "data/renter_households.rds")
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Havehead(quintile_raw, 7) they been measured by interviewing individuals or is the data coming from administrative sources?

# Part 3 - Quantifying

## 3.1 Data cleaning

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
# Variable 1: Affordability Ratio
# This shows how many years of income it takes to buy a home in each municipality
quintile_income <- quintile_income %>%
  mutate(affordability_ratio = avg_home_value / avg_income)

# check the result
quintile_income %>% select(municipality, avg_home_value, avg_income, affordability_ratio)
```

Variable 2

```{r gen_var2}
# Variable 2: Inequality Gap
# This shows the difference between high and low earners in each municipality
quintile_income <- quintile_income %>%
  mutate(inequality_gap = pct_top20 - pct_bottom40)

# check the result
quintile_income %>% select(municipality, pct_top20, pct_bottom40, inequality_gap)
```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
