---
title: "Income inequality and housing affordabilities in The Netherlands"
author: "Toms Kelderis (2898504)"
date: "`r Sys.Date()`"
output: pdf_document
---

#Lester M. \# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")

```

```{r packages}
require(tidyverse)li
require(rmarkdown)
require(yaml)
library(dplyr)
```

# Title Page

Toms Ķelderis (2898504), Nathan van Doorn (2801374), Frederick Nuboer (2809143), Khpelwaak Mamound (2786698), Caner Tepe (2785569), Mijke Gerritsen (2824547)

Tutorial 4, Group 1

Lecturer - Drs. Fatima Amankour

# Part 1 - Identify a Social Problem

Housing unaffordability and income inequality in the Netherlands

## 1.1 Describe the Social Problem

Text HERE

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and cleaning it}
dataset <- midwest

quintile_income <- read.csv("data/85039NED_Quintile_Income_2021.csv", sep = ";")

median_income <- read.csv("data/05064NED_Median_Income_2021.csv", sep = ";")

renter_households <- read.csv("data/82550NED_Renter_Owner_2021.csv", sep = ";")

#renaming quintile_income columns
quintile_income <- quintile_income %>%
  select(
    region_name        = WijkenEnBuurten,
    municipality       = Gemeentenaam_1,
    region_type        = SoortRegio_2,
    region_code        = Codering_3,
    avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
    pct_owner          = Koopwoningen_40,
    pct_renter         = HuurwoningenTotaal_41,
    pct_social_housing = InBezitWoningcorporatie_42,
    avg_income         = GemiddeldInkomenPerInwoner_72,
    pct_bottom40       = k_40PersonenMetLaagsteInkomen_73,
    pct_top20          = k_20PersonenMetHoogsteInkomen_74,
    pct_low_income     = HuishoudensMetEenLaagInkomen_78,
    pct_poverty_line   = HuishOnderOfRondSociaalMinimum_79 )

#filtering to municipalities only
quintile_income <- quintile_income %>%
  mutate(region_type = str_trim(region_type)) %>%
  filter(region_type == "Gemeente")

#fixing decimal commas
quintile_income <- quintile_income %>%
  mutate(across(c(avg_home_value, pct_owner, pct_renter, pct_social_housing,
                  avg_income, pct_bottom40, pct_top20, pct_low_income,
                  pct_poverty_line),
                ~ as.numeric(str_replace(., ",", "."))))

#renaming median_income columns
median_income <- median_income %>%
  select(
    population         = Populatie,
    household_type     = Kenmerken.van.huishoudens,
    year               = Perioden,
    region             = Regio.s,
    median_std_income  = Inkomen.Mediaan.gestandaardiseerd.inkomen..1.000.euro.,
    median_disp_income = Inkomen.Mediaan.besteedbaar.inkomen..1.000.euro. )

#renaming renter_households columns
renter_households <- renter_households %>%
  select(
    housing_type  = Woningtype,
    year          = Perioden,
    region        = Regio.s,
    housing_stock = Beginstand.woningvoorraad..aantal.)

saveRDS(quintile_income, "data/quintile_income.rds")
saveRDS(median_income, "data/median_income.rds")
saveRDS(renter_households, "data/renter_households.rds")
```

## 2.2 Provide a short summary of the dataset(s)

### Quintile Income

This dataset contains 13 selected variables across 20 Dutch municipalities, collected by CBS for the year 2021. Key variables include average home value in thousands of euros (based on WOZ tax assessment), average income per resident, and the share of residents in the bottom 40% and top 20% of national earners. A limitation is that the WOZ value is a tax assessment rather than an actual market price, and our selection covers only 20 out of approximately 340 Dutch municipalities.

### Median Income

This dataset contains 6 variables across 443 regions, collected by CBS for the year 2021. Key variables are median standardised income and median disposable income, both in thousands of euros, broken down by household type and region. We use this dataset at the national level for time series analysis.

### Renter Households

This dataset contains 4 variables across 64 regions, collected by CBS for the year 2021. The key variable is total housing stock broken down by housing type per region. We use this dataset at the national level to analyse differences between housing types.

## 2.3 Describe the type of variables included

### Quintile Income

The variables are socioeconomic (SES) in nature, covering income levels and housing ownership. The data comes from administrative sources — CBS combines tax records, municipal registers and housing registries rather than interviewing individuals.

### Median Income

The variables are SES in nature, focused on household income after taxes. The data comes from administrative tax records from the Dutch tax authority (Belastingdienst).

### Renter Households

The variables are structural rather than SES, they describe the physical housing supply rather than income. The data comes from administrative sources, specifically the Dutch national housing registry (BAG).

# Part 3 - Quantifying

## 3.1 Generate necessary variables

### Variable 1: Affordability Ratio

```{r gen_var1}
# Affordability ratio = average home value divided by average income (higher number = less affordable relative to income)
quintile_income <- quintile_income %>%
  mutate(affordability_ratio = avg_home_value / avg_income)

quintile_income %>% 
  select(municipality, avg_home_value, avg_income, affordability_ratio)

saveRDS(quintile_income %>% select(municipality, affordability_ratio),
        "data/affordability_ratio.rds")
```

### Variable 2: Inequality Gap

```{r gen_var2}
# Inequality gap = difference between top 20% and bottom 40% earners (Smaller the number = bigger gap between rich and poor)
quintile_income <- quintile_income %>%
  mutate(inequality_gap = pct_top20 - pct_bottom40)

quintile_income %>% 
  select(municipality, pct_top20, pct_bottom40, inequality_gap)

saveRDS(quintile_income %>% select(municipality, inequality_gap),
        "data/inequality_gap.rds")
```

## 3.2 Visualize temporal variation

```{r}

```

## 3.3 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.4 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
