---
title: "Income inequality and housing affordabilities in The Netherlands"
author: "Toms Kelderis (2898504)"
date: "`r Sys.Date()`"
output: pdf_document
---

#Lester M. \# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")
install.packages("cbsodataR")

```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(dplyr)
```

# Title Page

Toms Ķelderis (2898504), Nathan van Doorn (2801374), Frederick Nuboer (2809143), Khpelwaak Mamound (2786698), Caner Tepe (2785569), Mijke Gerritsen (2824547)

Tutorial 4, Group 1

Lecturer - Drs. Fatima Amankour

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?


1.What is the relationship between income inequality and housing affordability across Dutch municipalities?

Housing unaffordability and income inequality in the Netherlands
Housing affordability has become an increasingly important socioeconomic challenge in the Netherlands. Over the past decade, rising house prices and persistent housing shortages have reduced access to owner-occupied housing for many households. According to the Organisation for Economic Co-operation and Development (OECD, 2025), housing affordability remains a major policy concern, while the International Monetary Fund identifies insufficient housing supply as a key driver of affordability pressures in the Dutch housing market (Geis, 2023). 

At the same time, Dutch municipalities differ substantially in terms of income levels and income inequality. Previous research suggests that income inequality may influence housing outcomes through differences in purchasing power, housing market pressures, and access to housing opportunities (Dewilde & Lancee, 2012). As a result, municipalities with higher levels of income
inequality may experience different housing market dynamics than municipalities with more equal income distributions. 

Housing affordability is closely linked to financial security, wealth accumulation, and social mobility. Understanding whether municipalities with higher income inequality also experience lower housing affordability may therefore provide valuable insights into regional socioeconomic disparities. This study therefore examines the relationship between income inequality and housing affordability across Dutch municipalities, using the ratio of average housing values to income as an indicator of affordability.


# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and cleaning it}
library(tidyverse)
library(cbsodataR)

dataset <- midwest

# Loading in all datasets through CBS API 

# quintile_income - one dataset per year
kwb_2015 <- cbs_get_data("83220NED")
kwb_2016 <- cbs_get_data("83487NED")
kwb_2017 <- cbs_get_data("83765NED")
kwb_2018 <- cbs_get_data("84286NED")
kwb_2019 <- cbs_get_data("84583NED")
kwb_2020 <- cbs_get_data("84799NED")
kwb_2021 <- cbs_get_data("85039NED")

# median_income -> single dataset with all years
median_income_raw <- cbs_get_data("85064NED")

# renter_households -> single dataset with all years
renter_households_raw <- cbs_get_data("82550NED")

# Cleaning quintile_income dataset 

# 2015 only -> different home value name AND income columns
clean_kwb_2015 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWoningwaarde_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_66,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
      pct_top20          = k_20PersonenMetHoogsteInkomen_68,
      pct_low_income     = HuishoudensMetEenLaagInkomen_72,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73
    ) %>%
    mutate(year = yr)
}

# 2016 and 2017 -> same income columns as 2015 but different home value name
clean_kwb_1617 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_66,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
      pct_top20          = k_20PersonenMetHoogsteInkomen_68,
      pct_low_income     = HuishoudensMetEenLaagInkomen_72,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73
    ) %>%
    mutate(year = yr)
}

# 2018 only -> different income column numbers
clean_kwb_2018 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_codey        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_69,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_70,
      pct_top20          = k_20PersonenMetHoogsteInkomen_71,
      pct_low_income     = HuishoudensMetEenLaagInkomen_75,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_76
    ) %>%
    mutate(year = yr)
}

# 2019, 2020 and 2021 -> same column numbers
clean_kwb_1921 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_72,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_73,
      pct_top20          = k_20PersonenMetHoogsteInkomen_74,
      pct_low_income     = HuishoudensMetEenLaagInkomen_78,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_79
    ) %>%
    mutate(year = yr)
}

# mergeing all years
quintile_income <- bind_rows(
  clean_kwb_2015(kwb_2015, 2015),
  clean_kwb_1617(kwb_2016, 2016),
  clean_kwb_1617(kwb_2017, 2017),
  clean_kwb_2018(kwb_2018, 2018),
  clean_kwb_1921(kwb_2019, 2019),
  clean_kwb_1921(kwb_2020, 2020),
  clean_kwb_1921(kwb_2021, 2021))

# fixing decimal commas
quintile_income <- quintile_income %>%
  mutate(across(c(avg_home_value, pct_owner, pct_renter, pct_social_housing,
                  avg_income, pct_bottom40, pct_top20, pct_low_income,
                  pct_poverty_line),
                ~ as.numeric(str_replace(as.character(.), ",", "."))))

# Cleaning median_income dataset 

median_income <- median_income_raw %>%
  select(
    population         = Populatie,
    household_type     = KenmerkenVanHuishoudens,
    year               = Perioden,
    region             = RegioS,
    median_std_income  = MediaanGestandaardiseerdInkomen_4,
    median_disp_income = MediaanBesteedbaarInkomen_6)

median_income <- median_income %>%
  mutate(year = as.integer(str_sub(year, 1, 4)))

median_income <- median_income %>%
  filter(year >= 2015 & year <= 2020)

unique(median_income$year)

# Cleaning renter_households dataset

renter_households <- renter_households_raw %>%
  select(
    housing_type  = Woningtype,
    year          = Perioden,
    region        = RegioS,
    housing_stock = BeginstandWoningvoorraad_1 )

renter_households <- renter_households %>%
  mutate(
    year = as.integer(str_sub(year, 1, 4)),
    region = str_trim(region) )

renter_households <- renter_households %>%
  mutate(housing_type = case_when(
    housing_type == "T001100" ~ "Total",
    housing_type == "ZW10290" ~ "Owner-occupied",
    housing_type == "ZW10340" ~ "Rental",
    TRUE ~ housing_type))

renter_households <- renter_households %>%
  filter(year >= 2015 & year <= 2021)

unique(renter_households$year)

# Saving cleaned datasets to Data folder

saveRDS(quintile_income,   "data/quintile_income.rds")
saveRDS(median_income,     "data/median_income.rds")
saveRDS(renter_households, "data/renter_households.rds")
```

## 2.2 Provide a short summary of the dataset(s)

### Quintile Income

This dataset contains 15 variables across 373 Dutch municipalities, collected by CBS covering 2015 to 2021. Key variables include average home value (WOZ tax assessment), average income per resident, income distribution indicators, and two newly constructed variables: affordability ratio and inequality gap. A limitation is that WOZ value may underestimate true market prices and municipality boundaries changed slightly between years.

### Median Income

This dataset contains 6 variables across Dutch regions, collected by CBS covering 2015 to 2020. Key variables are median standardised and disposable income in thousands of euros, broken down by household type. We use this for national-level time series analysis. A limitation is that 2021 data is not yet available.

### Renter Households

This dataset contains 4 variables across Dutch regions, collected by CBS covering 2015 to 2021. The key variable is total housing stock broken down into total, owner-occupied, and rental. A limitation is that it contains no price or income information.

## 2.3 Describe the type of variables included

### Quintile Income

The variables are socioeconomic (SES) in nature, covering income levels, housing ownership, poverty indicators, and two newly constructed analytical variables. The data comes from administrative sources — CBS combines tax records, municipal registers and housing registries rather than interviewing individuals.

### Median Income

The variables are SES in nature, focused on household income after taxes broken down by household type. The data comes from administrative tax records from the Dutch tax authority (Belastingdienst).

### Renter Households

The variables are structural rather than SES — they describe the physical housing supply by tenure type. The data comes from administrative sources, specifically the Dutch national housing registry (BAG).

# Part 3 - Quantifying

## 3.1 Generate necessary variables

### Variable 1: Affordability Ratio

```{r gen_var1}
# affordability ratio = home value divided by income
# higher number = less affordable
quintile_income <- quintile_income %>%
  mutate(affordability_ratio = avg_home_value / avg_income)

# check
quintile_income %>%
  select(municipality, year, avg_home_value, avg_income, affordability_ratio) %>%
  head()

# Saveing updated quintile_income with affordability_ratio
saveRDS(quintile_income, "data/quintile_income.rds")

```

### Variable 2: Inequality Gap

```{r gen_var2}
# inequality gap = difference between top 20% and bottom 40% earners
# more negative = more unequal
quintile_income <- quintile_income %>%
  mutate(inequality_gap = pct_top20 - pct_bottom40)

# check
print(quintile_income %>%
  select(municipality, year, pct_top20, pct_bottom40, inequality_gap) %>%
  head())

# Save updated quintile_income with both new variables
saveRDS(quintile_income, "data/quintile_income.rds")
```

## 3.2 Visualize temporal variation

```{r}

```

## 3.3 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.4 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
