---
title: "Income inequality and housing affordabilities in The Netherlands"
author: "Toms Kelderis (2898504)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

#Lester M. \# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(dplyr)
```

# Title Page

Toms Ķelderis (2898504), Nathan van Doorn (2801374), Frederick Nuboer (2809143), Khpelwaak Mamound (2786698), Caner Tepe (2785569), Mijke Gerritsen (2824547)

Tutorial 4, Group 1

Lecturer - Drs. Fatima Amankour

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?


1.What is the relationship between income inequality and housing affordability across Dutch municipalities?

Housing unaffordability and income inequality in the Netherlands
Housing affordability has become an increasingly important socioeconomic challenge in the Netherlands. Over the past decade, rising house prices and persistent housing shortages have reduced access to owner-occupied housing for many households. According to the Organisation for Economic Co-operation and Development (OECD, 2025), housing affordability remains a major policy concern, while the International Monetary Fund identifies insufficient housing supply as a key driver of affordability pressures in the Dutch housing market (Geis, 2023). 

At the same time, Dutch municipalities differ substantially in terms of income levels and income inequality. Previous research suggests that income inequality may influence housing outcomes through differences in purchasing power, housing market pressures, and access to housing opportunities (Dewilde & Lancee, 2012). As a result, municipalities with higher levels of income
inequality may experience different housing market dynamics than municipalities with more equal income distributions. 

Housing affordability is closely linked to financial security, wealth accumulation, and social mobility. Understanding whether municipalities with higher income inequality also experience lower housing affordability may therefore provide valuable insights into regional socioeconomic disparities. This study therefore examines the relationship between income inequality and housing affordability across Dutch municipalities, using the ratio of average housing values to income as an indicator of affordability.


# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and cleaning it}
library(tidyverse)
library(cbsodataR)

dataset <- midwest

# Loading in all datasets through CBS API 

# quintile_income - one dataset per year
kwb_2015 <- cbs_get_data("83220NED")
kwb_2016 <- cbs_get_data("83487NED")
kwb_2017 <- cbs_get_data("83765NED")
kwb_2018 <- cbs_get_data("84286NED")
kwb_2019 <- cbs_get_data("84583NED")
kwb_2020 <- cbs_get_data("84799NED")
kwb_2021 <- cbs_get_data("85039NED")

# median_income -> single dataset with all years
median_income_raw <- cbs_get_data("85064NED")

# renter_households -> single dataset with all years
renter_households_raw <- cbs_get_data("82550NED")

# Cleaning quintile_income dataset 

# 2015 only -> different home value name AND income columns
clean_kwb_2015 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWoningwaarde_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_66,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
      pct_top20          = k_20PersonenMetHoogsteInkomen_68,
      pct_low_income     = HuishoudensMetEenLaagInkomen_72,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73
    ) %>%
    mutate(year = yr)
}

# 2016 and 2017 -> same income columns as 2015 but different home value name
clean_kwb_1617 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_66,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_67,
      pct_top20          = k_20PersonenMetHoogsteInkomen_68,
      pct_low_income     = HuishoudensMetEenLaagInkomen_72,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_73
    ) %>%
    mutate(year = yr)
}

# 2018 only -> different income column numbers
clean_kwb_2018 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_codey        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_69,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_70,
      pct_top20          = k_20PersonenMetHoogsteInkomen_71,
      pct_low_income     = HuishoudensMetEenLaagInkomen_75,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_76
    ) %>%
    mutate(year = yr)
}

# 2019, 2020 and 2021 -> same column numbers
clean_kwb_1921 <- function(df, yr) {
  df %>%
    mutate(SoortRegio_2 = str_trim(SoortRegio_2)) %>%
    filter(SoortRegio_2 == "Gemeente") %>%
    select(
      region_name        = WijkenEnBuurten,
      municipality       = Gemeentenaam_1,
      region_type        = SoortRegio_2,
      region_code        = Codering_3,
      avg_home_value     = GemiddeldeWOZWaardeVanWoningen_35,
      pct_owner          = Koopwoningen_40,
      pct_renter         = HuurwoningenTotaal_41,
      pct_social_housing = InBezitWoningcorporatie_42,
      avg_income         = GemiddeldInkomenPerInwoner_72,
      pct_bottom40       = k_40PersonenMetLaagsteInkomen_73,
      pct_top20          = k_20PersonenMetHoogsteInkomen_74,
      pct_low_income     = HuishoudensMetEenLaagInkomen_78,
      pct_poverty_line   = HuishOnderOfRondSociaalMinimum_79
    ) %>%
    mutate(year = yr)
}

# mergeing all years
quintile_income <- bind_rows(
  clean_kwb_2015(kwb_2015, 2015),
  clean_kwb_1617(kwb_2016, 2016),
  clean_kwb_1617(kwb_2017, 2017),
  clean_kwb_2018(kwb_2018, 2018),
  clean_kwb_1921(kwb_2019, 2019),
  clean_kwb_1921(kwb_2020, 2020),
  clean_kwb_1921(kwb_2021, 2021))

# fixing decimal commas
quintile_income <- quintile_income %>%
  mutate(across(c(avg_home_value, pct_owner, pct_renter, pct_social_housing,
                  avg_income, pct_bottom40, pct_top20, pct_low_income,
                  pct_poverty_line),
                ~ as.numeric(str_replace(as.character(.), ",", "."))))

# Cleaning median_income dataset 

median_income <- median_income_raw %>%
  select(
    population         = Populatie,
    household_type     = KenmerkenVanHuishoudens,
    year               = Perioden,
    region             = RegioS,
    median_std_income  = MediaanGestandaardiseerdInkomen_4,
    median_disp_income = MediaanBesteedbaarInkomen_6)

median_income <- median_income %>%
  mutate(year = as.integer(str_sub(year, 1, 4)))

median_income <- median_income %>%
  filter(year >= 2015 & year <= 2020)

unique(median_income$year)

# Cleaning renter_households dataset

renter_households <- renter_households_raw %>%
  select(
    housing_type  = Woningtype,
    year          = Perioden,
    region        = RegioS,
    housing_stock = BeginstandWoningvoorraad_1 )

renter_households <- renter_households %>%
  mutate(
    year = as.integer(str_sub(year, 1, 4)),
    region = str_trim(region) )

renter_households <- renter_households %>%
  mutate(housing_type = case_when(
    housing_type == "T001100" ~ "Total",
    housing_type == "ZW10290" ~ "Owner-occupied",
    housing_type == "ZW10340" ~ "Rental",
    TRUE ~ housing_type))

renter_households <- renter_households %>%
  filter(year >= 2015 & year <= 2021)

unique(renter_households$year)

# Saving cleaned datasets to Data folder

saveRDS(quintile_income,   "data/quintile_income.rds")
saveRDS(median_income,     "data/median_income.rds")
saveRDS(renter_households, "data/renter_households.rds")
```

## 2.2 Provide a short summary of the dataset(s)

### Quintile Income

This dataset contains 15 variables across 373 Dutch municipalities, collected by CBS covering 2015 to 2021. Key variables include average home value (WOZ tax assessment), average income per resident, income distribution indicators, and two newly constructed variables: affordability ratio and inequality gap. A limitation is that WOZ value may underestimate true market prices and municipality boundaries changed slightly between years.

### Median Income

This dataset contains 6 variables across Dutch regions, collected by CBS covering 2015 to 2020. Key variables are median standardised and disposable income in thousands of euros, broken down by household type. We use this for national-level time series analysis. A limitation is that 2021 data is not yet available.

### Renter Households

This dataset contains 4 variables across Dutch regions, collected by CBS covering 2015 to 2021. The key variable is total housing stock broken down into total, owner-occupied, and rental. A limitation is that it contains no price or income information.

## 2.3 Describe the type of variables included

### Quintile Income

The variables are socioeconomic (SES) in nature, covering income levels, housing ownership, poverty indicators, and two newly constructed analytical variables. The data comes from administrative sources — CBS combines tax records, municipal registers and housing registries rather than interviewing individuals.

### Median Income

The variables are SES in nature, focused on household income after taxes broken down by household type. The data comes from administrative tax records from the Dutch tax authority (Belastingdienst).

### Renter Households

The variables are structural rather than SES — they describe the physical housing supply by tenure type. The data comes from administrative sources, specifically the Dutch national housing registry (BAG).

# Part 3 - Quantifying

## 3.1 Generate necessary variables

### Variable 1: Affordability Ratio

```{r gen_var1}
# affordability ratio = home value divided by income
# higher number = less affordable
quintile_income <- quintile_income %>%
  mutate(affordability_ratio = avg_home_value / avg_income)

# check
quintile_income %>%
  select(municipality, year, avg_home_value, avg_income, affordability_ratio) %>%
  head()

# Saveing updated quintile_income with affordability_ratio
saveRDS(quintile_income, "data/quintile_income.rds")

```

### Variable 2: Inequality Gap

```{r gen_var2}
# inequality gap = difference between top 20% and bottom 40% earners
# more negative = more unequal
quintile_income <- quintile_income %>%
  mutate(inequality_gap = pct_top20 - pct_bottom40)

# check
print(quintile_income %>%
  select(municipality, year, pct_top20, pct_bottom40, inequality_gap) %>%
  head())

# Save updated quintile_income with both new variables
saveRDS(quintile_income, "data/quintile_income.rds")
```

## 3.2 Visualize temporal variation

```{r}
income_plot_data <- filter(median_income, region == "NL01  ", household_type == "1050010", population == "1050010")

ggplot(income_plot_data, aes(x = year, y = median_disp_income)) +
  geom_line() +
  geom_point() +
  labs(title = "Median disposable household income in the Netherlands, 2015-2020",
       x = "Year", y = "Median disposable income (x1,000 euros)",
       caption = "Source: CBS (85064NED)") +
  theme_minimal() 


housing_plot_data <- filter(renter_households, region == "NL01", housing_type %in% c("Total", "Owner-occupied", "Rental"))

ggplot(housing_plot_data, aes(x = year, y = housing_stock, color = housing_type)) +
  geom_line() +
  geom_point() +
  labs(title = "Housing stock by tenure type in the Netherlands, 2015-2021",
       x = "Year", y = "Number of dwellings", color = "Housing type",
       caption = "Source: CBS (82550NED)") +
  theme_minimal() 
```

Figure 3.2.1 shows that median disposable household income in the Netherlands increased steadily from €32,500 in 2015 to €38,000 in 2020, an increase of approximately 16.9% over the six-year period (roughly 2.7% per year on average). Growth was relatively modest in the earlier years — around €0.7-0.9 thousand per year between 2015 and 2018 — but accelerated noticeably from 2018 onwards, with increases of €1.7 thousand (2018-2019) and €1.3 thousand (2019-2020). This indicates that income growth gained momentum towards the end of the period rather than following a perfectly linear trend. Since the affordability ratio used elsewhere in this report (Section 3.1) is defined as average home value divided by average income, this income series represents the denominator of that ratio: rising incomes work to improve affordability in isolation, but whether affordability actually improved over this period depends on how housing values developed over the same years.

Figure 3.2.2 shows the development of the Dutch housing stock by tenure type between 2015 and 2021. The total housing stock grew from approximately 7.59 million dwellings in 2015 to 7.97 million in 2021, an increase of about 378,000 dwellings (+5.0%). Both tenure types contributed similar absolute increases — owner-occupied housing grew from roughly 4.91 million to 5.10 million dwellings (+189,800, +3.9%), while the rental segment grew from about 2.68 million to 2.87 million dwellings (+188,500, +7.0%). Because the rental segment started from a smaller base, its relative growth rate was nearly double that of owner-occupied housing, so the rental share of the total stock rose slightly, from 35.3% in 2015 to 36.0% in 2021. Overall, the expansion of the housing stock was gradual and fairly steady across the period, without sharp jumps or declines. Comparing the two figures, median income grew by roughly 17% between 2015 and 2020, considerably faster than the 5% growth in total housing stock over a comparable period (2015-2021). If housing demand scales with rising incomes and population growth, this relatively slow expansion of supply — particularly in the owner-occupied segment — may help explain the affordability pressures identified in the spatial analysis (Section 3.3), where municipalities with higher average incomes also tend to show the highest affordability ratios.
## 3.3 Visualize spatial variation

```{r visualise_map, fig.cap="Housing affordability ratio across Dutch municipalities in 2021."}

library(sf)
library(dplyr)
library(ggplot2)
library(viridisLite)

# JOINING MUNICIPALITY BOUNDARIES DATA WITH AFFORDABILITY DATA:
map_data<-quintile_income%>%
  filter(year==2021)%>%
  mutate(region_code=trimws(region_code))

municipalities_map<-municipalities%>%
  left_join(map_data,
            by=c("statcode"="region_code"))

# MAKING THE MAP
library(ggplot2)

ggplot(municipalities_map)+
  geom_sf(aes(fill=affordability_ratio))+
  scale_fill_viridis_c()+ # default appearance
  labs(
    title = "Housing Affordability Across Dutch Municipalities (2021)",
    subtitle = "Affordability ratio by munipality",
    fill = "Affordability\nRatio",
    caption = "Source: CBS (2021)"
  )+
  theme_minimal()
# changing appearance: scale_fill_viridis_c(option = "plasma" OR "magma" OR "cividis")

# Top 10 municipalities with "worst" affordability (for stronger interpretation)
municipalities_map %>%
  st_drop_geometry() %>%
  arrange(desc(affordability_ratio)) %>%
  select(statnaam, affordability_ratio) %>%
  head(10)
```

Figure 3.3 presents the housing affordability ratio across Dutch municipalities in 2021. The affordability ratio is calculated as average home value divided by average income, where higher values indicate lower affordability because housing prices are high relative to local incomes.

The map reveals substantial spatial variation in housing affordability across the Netherlands. Several municipalities display particularly high affordability ratios, including Bloemendaal (16.35), Blaricum (15.95), Laren (15.46), Landsmeer (13.70), and Wassenaar (13.36). These municipalities experience greater affordability pressures because house values are high relative to local incomes. In contrast, many other municipalities exhibit lower affordability ratios, indicating that housing is more affordable relative to income.

This figure is directly relevant to the social problem of housing affordability because it highlights geographic inequalities in access to housing. The results show that affordability pressures are not evenly distributed across the country, with some municipalities facing substantially greater challenges than others. Understanding these spatial differences is important for policymakers, as local housing markets may require different policy responses than those suggested by national averages.

## 3.4 Visualize sub-population variation

```{r subpop_boxplot}
# I made both grp'ed barchart and boxplot. I think boxplot is more suited to our investigation but grouped barchart is mentioned in the rubric if not mistaken. 

# filtering
subpop_data<-quintile_income%>%
  filter(year==2021)

# income categories (groups)
subpop_data<-subpop_data%>%
  mutate(
    income_category=case_when(
      avg_income<25~"Low income",
      avg_income<=32~"Medium income",
      TRUE~"High income"
    )
  )

# renter groups
subpop_data<-subpop_data%>%
  mutate(
    renter_group=case_when(
      pct_renter>40~"High renter share",
      TRUE~"Low renter share"
    )
  )

# BOXPLOT
ggplot(
  subpop_data,
  aes(
    x=income_category,
    y=affordability_ratio,
    fill=renter_group
  )
)+
  geom_boxplot()+
  labs(
    title="Housing affordability by income category and renter share (2021).",
    x="Income category",
    y="Affordability ratio",
    fill="Renter share",
    caption="Source: CBS (2021)"
  )+
  scale_fill_viridis_d()+
  theme_minimal()
# calculating means (for interpretation)
subpop_data%>%
  group_by(income_category,renter_group)%>%
  summarise(
    mean_affordability=mean(affordability_ratio, na.rm=TRUE),
    n=n()
  )
```

Figure 3.4.1 shows how housing affordability varies across municipalities grouped by income category and renter share in 2021. A clear pattern emerges: municipalities in the high-income category have the highest affordability ratios, while municipalities in the low-income category have the lowest. Because the affordability ratio is calculated as average home value divided by average income, a higher ratio indicates that housing is less affordable relative to local incomes. This suggests that rising housing values in affluent municipalities more than offset their higher incomes. Within both the low- and medium-income categories, municipalities with lower renter shares tend to have higher affordability ratios than municipalities with higher renter shares, indicating slightly lower affordability. The figure is relevant to the housing affordability problem because it demonstrates that affordability pressures are not evenly distributed across municipalities and are particularly pronounced in wealthier areas with expensive housing markets.

```{r subpop_groupedbarchart}
# summary dataset
bar_data<-subpop_data%>%
  group_by(income_category,renter_group)%>%
  summarise(
    mean_affordability=
      mean(affordability_ratio,na.rm=TRUE)
  )

# PLOTTING
ggplot(
  bar_data,
  aes(
    x=income_category,
    y=mean_affordability,
    fill=renter_group
  )
)+
  geom_col(position = "dodge")+
  labs(
    title="Average housing affordability by income category and renter share (2021)",
    x="Income category",
    y="Average affordability ratio",
    fill="Renter share",
    caption ="Source: CBS (2021)"
  )+
  scale_fill_viridis_d()+
  theme_minimal()
```
Figure 3.4.1 compares the average affordability ratio across municipalities grouped by income category and renter share in 2021. Average affordability ratios increase steadily from low-income to high-income municipalities, indicating that housing becomes less affordable relative to income as average incomes rise. In all income categories, municipalities with lower renter shares have slightly higher affordability ratios than municipalities with higher renter shares. This suggests that municipalities with larger rental sectors may experience somewhat lower affordability pressures than municipalities dominated by owner-occupied housing. The figure is relevant to the social problem because it highlights how housing affordability differs across municipal sub-populations, helping identify which types of municipalities face the greatest challenges in accessing affordable housing.

## 3.5 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
